---
title: "Using gp to wrangle plate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gp to wrangle plate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gp)
```

# Introduction

If you work in a lab or with lab data, you've probably had to deal with data that look like this:

```{r}
pcr_plate
```

Just looking at this sucks. You can use some `tidyverse` magic or use `plater` to help you at least make this matrix into a tidy data.frame, but annotating the wells as to which rows and which columns are what can be soul-rending (heavens forbid you have a non-standard plate layout). In this vignette, we'll not only tidy these data, but annotate them in a fairly painless process.


# The data

The data shown above represent the relative quantities (RQs) of a qPCR run with three primers:

```{r, echo = FALSE}
gp(16, 24) |> 
  gp_sec("no_border", margin = 1) |> 
  gp_sec("primers", nrow = 9, ncol = 3, c("Primer 1", "Primer 2", "Primer 3")) |> 
  gp_plot(primers)

```

Four cell-lines, plus a non-targetting control (NTC):

```{r}
gp(16, 24) |> 
  gp_sec("no_border", margin = 1) |> 
  gp_sec("primers", nrow = 9, ncol = 3, labels = c("Primer 1", "Primer 2", "Primer 3")) |> 
  gp_sec("cell line", nrow = 2, labels = c("1", "2", "3", "4")) |> 
  gp_plot(`cell line`)
```

